---
title:  "Pipeline for least square post-imputation inference for proteomic data"
author: "Haeun Moon"
output: html_document
date: "2024-03-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache= FALSE)
```

# Introduction

This notebook illustrate how to perform a post-imputation inference of proteomic data with missing values on example dataset. By replacing the example dataset with the dataset of your choice, you can draw an inference on least square coefficients on your own data.

For more information and details on the theoretical background of the methodology, we refer to the paper [[@mayer2020]](http://dx.doi.org/10.1214/20-AOAS1356).

**You can replace the chunk `rename_data` with any data set you wish to analyze. 
For this, you have to specify:**

- `raw.pep`: high-dimensional peptide data outcome with missing values. (A matrix #observations x #peptides)
- `covariate`: A low-dimensional covariate without missing values. (A data frame #observations x #covariates)
- `missing_pattern`: An assumed missing pattern of `raw.pep`. Either `MAR` or `MCAR`.


# Preliminaries

## Load libraries

```{r load_libraries, results='hide', message=F, warning=F}
require(mvtnorm)
library(stats)
require(sandwich)
require(clusterGeneration)

source("R_wrapper_VAE.R")

```

## Generate example dataset

Two data are generated: `raw.pep` consists of high-dimensional peptide abundance data with an arbitrary correlation structure and MAR missing patterns, and `covariate` consists of low-dimensional covariate information data, which is fully observed.

```{r choose_parameters}
n <- 500 # number of observations
p <- 1000 # dimension of peptide data 
e <- 0.3 # parameter for missing rate 
set.seed(123)
```


```{r generate_toy_data}

# Low dimensional covariate with dimension 2 are generated
W1=runif(n,0,1) # A covariate of interest
W2=runif(n,0,1) # Other covariate(s) 
covariate=data.frame(W1,W2) 

#High-dimensional outcome matrix without missingness is generated from an arbitrary correlation structure 
C0 <- rcorrmatrix(p/10, alphad = 1) 
C<-diag(10) %x% C0
comp.pep=rmvnorm(n, mean=rep(0,p),sigma=C)+replicate(p,covariate$W2)

#Signals are injecting to true peptides upon a random selection of 10% of total peptides.
true.peptide=sample(p, p*0.1)
comp.pep[,true.peptide]=comp.pep[,true.peptide]+0.6*replicate(length(true.peptide),covariate$W1) 

#Missing masks are generated under the MAR assumption. A rate of missingness is roughly around a tuning parameter 'e'. If a certain peptide has an excessively high rate of missingness, imputation models will not work satisfactorily. Therefore, we impose a safety cap on the missing rate of individual peptides (set at 0.8).

ex=exp(covariate$W2)/(1+exp(covariate$W2))/2
misp=rbinom(n, p, prob=ifelse(ex*e/mean(ex)>0.8, 0.8, ex*e/mean(ex)))

generate_sample <- function(size) {sample(p, size = size, replace = FALSE)}
ij.na=cbind(rep(c(1:n), misp),unlist(lapply(misp, generate_sample)))

# Peptide data with missingness is generated.   
raw.pep=comp.pep
raw.pep[ij.na]=NA

toy_data <- list(raw.pep=raw.pep, covariate=covariate, missing_pattern="MAR")
```

**If you want to use your own data set, change the following chunk**

```{r rename_data}
raw.pep <- toy_data$raw.pep
covariate <- data.frame(toy_data$covariate)
missing_pattern<-toy_data$missing_pattern

n <- dim(raw.pep)[1]
p <- dim(raw.pep)[2]
```

# Imputation

We fit an outcome model using both low-dimensional covariates and the entire peptide matrix. A challenge is that a peptide matrix has many missing entries even when used as a covariate in the regression problem. Our employed model is a variant of variational autoencoder, called VAEIT, a deep neural network tool that allows for flexible input and simultaneous estimation of the multi-response regression.

The VAEIT model accepts input for a low-dimensional covariate in two types: continuous (batches_cont) and categorical (batches_cate). **If you are using your own dataset, and if it contains categorical low-dimensional covariates, adjust batches_cate and batches_cont accordingly.**

```{r fitting_outcome}

batches_cate=NULL
batches_cont=cbind(covariate$W1, covariate$W2)
        
data <-raw.pep; data[is.na(data)] <- 0.;
mask <- - as.matrix(is.na(as.matrix(raw.pep)))
vae.impute<-vae(data,mask,batches_cate,batches_cont, NULL,dist_block=list('Gaussian'),
                        c(dim(data)[2]), num_epoch=300L, verbose=FALSE)[[1]]


```

# Construct pseudo-outcomes

Pseudo-outcomes are constructed through a predefined formula which utilize fitted outcomes and propensity scores. Fitted outcomes were obtained from a previous chunk. We further fit propensity scores using a logit model.

```{r construct pseudo-outcomes, message = F, warning = F}

fitd=raw.pep

  for(i.p in 1:p){
    mask=1-is.na(raw.pep[,i.p])
    prop=glm(mask~.,data=covariate, family = "binomial")
    fitd[,i.p]<-as.vector(prop$fitted.values)
  }

  AG=(raw.pep-vae.impute); AG[which(is.na(AG),arr.ind = T)]=0
  DR.pep=vae.impute+AG/fitd 

```


# Linear regression analysis

We perform a linear regression for peptide abundance on  on low-dimensional covariates. As a response, each column of the doubly robust psuedo-outcomes `DR.pep` (proposed method), or the fitted outcomes `vae.impute` (plugin method) are considered. The covariate of interest is set to the first column of the data frame `covariate`, but it is subject to change for analysis purposes.

A choice of variance estimator for the t-test depends on an missing pattern of  `raw.pep` data. Under the MAR missing pattern, a heteroskedastic-consistent estimator is used, and under the assumption of Missing Completely at Random (MCAR), this simplifies to using the usual Ordinary Least Squares (OLS) estimator.

```{r linear_regression, message = F, warning = F}


pval=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))
tval=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))
beta=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))

if(missing_pattern=="MAR"){
  for(i.p in 1:p){
    fit=lm(DR.pep[,i.p]~., data=covariate)
    beta$DR_UW[i.p]=coef(summary(fit))[2,1] 
    tval$DR_UW[i.p]=beta$DR_UW[i.p]/sqrt(vcovHC(fit)[2,2])
    pval$DR_UW[i.p]=2*pt(abs(tval$DR_UW[i.p]), n-1, lower.tail=FALSE) 
    
    fit=lm(vae.impute[,i.p]~., data=covariate)
    pval$plugin[i.p]=coef(summary(fit))[2,4] 
    tval$plugin[i.p]=coef(summary(fit))[2,3] 
    beta$plugin[i.p]=coef(summary(fit))[2,1] 
    }
} else if(missing_pattern=="MCAR"){
    for(i.p in 1:p){
    fit=lm(DR.pep[,i.p]~., data=covariate)
    beta$DR_UW[i.p]=coef(summary(fit))[2,1] 
    tval$DR_UW[i.p]=coef(summary(fit))[2,3] 
    pval$DR_UW[i.p]=coef(summary(fit))[2,4] 
    
    fit=lm(vae.impute[,i.p]~W1+W2, data=covariate)
    pval$plugin[i.p]=coef(summary(fit))[2,4] 
    tval$plugin[i.p]=coef(summary(fit))[2,3] 
    beta$plugin[i.p]=coef(summary(fit))[2,1] 
    }
  } else{print('Invalid missing pattern')}
  

```

# Multiple Testing procedure


We combine the p-values derived in `linear regression` with a multiple testing procedure, to make discoveries of important peptides associated with a covariate of interest.

Benjamini-Hochberg procedure [[@benjamini1995controlling]] is applied for a multiple testing. First, we transform the p-values to q-values, and identify the indices whose q-values are less than a predefined cutoff `tFDR`. A value of`tFDR` can be set according to the purpose of analysis. Typical values include 0.01, 0.05, 0.1, or 0.3.

```{r multiple_testing}

qval=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))
qval$DR_UW=p.adjust(pval$DR_UW, method = "BH")
qval$plugin=p.adjust(pval$plugin, method = "BH")

tFDR=0.3
which(qval$DR_UW<tFDR);

```

## Evaluation of methods

When synthetics were used, we can evaluate the algorithm based on a known ground truth of truly associated peptides (indices in "true.peptide"). We evaluate the performance of the algorithm based on two criteria; 1) the fraction of false discoveries over the number of total discoveries (FDR) and 2) the fraction of true discoveries over the number of true peptides (power). We compare the performance of the algorithm with the case when the imputed values are directly plugged in for pseudo-outcomes ("Plugin method". An ideal method has well-controlled FDR and a power close to one. 

```{r evaluation}

power_DR=length(intersect(which(qval$DR_UW<tFDR),true.peptide))/length(true.peptide)
FDR_DR=length(setdiff(which(qval$DR_UW<tFDR),true.peptide))/length(which(qval$DR_UW<tFDR))

print(paste0("FDR (proposed) : ", round(FDR_DR,3),", Power (proposed) : ", round(power_DR,3) ))

power_plugin=length(intersect(which(qval$plugin<tFDR),true.peptide))/length(true.peptide)
FDR_plugin=length(setdiff(which(qval$plugin<tFDR),true.peptide))/length(which(qval$plugin<tFDR))


print(paste0("FDR (plug-in) : ", round(FDR_plugin,3), ", Power (plug-in) : ", round(power_plugin,3)))



```

# References
