---
title: "pipeline_toy"
output: html_document
date: "2024-03-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache= TRUE)
```

# Introduction

This notebook illustrate how to perform a post-imputation inference of proteomic data with missing values on example dataset. By replacing the example dataset with the dataset of your choice, you can draw an inference on least square coefficients on your own data.

For more information and details on the theoretical background of the methodology, we refer to the paper [[@mayer2020]](http://dx.doi.org/10.1214/20-AOAS1356).

**You can replace the chunks `rename_data` and `confounders` with any data set you wish to analyze. 
For this, you have to specify:**

- `W`: low-dimensional covariates without missing values.
- `raw.pep``: high-dimensional peptide data outcome with missing values.


# Preliminaries

## Load libraries

```{r load_libraries, results='hide', message=F, warning=F}
require(data.table)
require(mvtnorm)
library(stats)
require(sandwich)
library(ggplot2)
require(dplyr)
require(clusterGeneration)

source("0. VAE_python.R")

set.seed(1234)
```

## Generate toy example

We will generate a simple toy dataset with normally distributed low-dimensional covariates and high dimensional peptide data with MCAR missing patterns.
```{r choose_parameters}
seed <- 4321
n <- 500 # number of observations
p <- 1000 # dimension of peptide data 
e <- 0.3 # parameter for missing rate 
```


```{r generate_toy_data}
# synthetic data is generated for illustration purpose.
# Generating low dimensional covariates, with dimension 2
# Set a first column data of covariate dataframe as a variable of interest

W1=runif(n,0,1) # low-dimensional covariateof interest
W2=runif(n,0,1) # low-dimensional covariate 
covariate=data.frame(W1,W2)

#High-dimensional outcome matrix without missingness is generated 
#from an arbitrary correlation structure 

C0 <- rcorrmatrix(p/10, alphad = 1) 
C<-diag(10) %x% C0
comp.pep=rmvnorm(n, mean=rep(0,p),sigma=C)+replicate(p,W2)

#injecting signals to true peptides (10% of total number p)

true.peptide=sample(p, p*0.1) 
comp.pep[,true.peptide]=comp.pep[,true.peptide]+0.5*replicate(length(true.peptide),W1) 

#generating missing mask 

ex=exp(W2)/(1+exp(W2))/2
misp=rbinom(n, p, prob=ifelse(ex*e/mean(ex)>0.8, 0.8, ex*e/mean(ex)))

generate_sample <- function(size) {sample(p, size = size, replace = FALSE)}
ij.na=cbind(rep(c(1:n), misp),unlist(lapply(misp, generate_sample)))

# Generating outcome with missingness
    
raw.pep=comp.pep
raw.pep[ij.na]=NA

toy_data <- list(raw.pep=raw.pep, covariate=covariate)
```

**If you want to use your own data set, change the following chunk**

```{r rename_data}
raw.pep <- data.frame(toy_data$raw.pep)
covariate <- toy_data$covariate

n <- dim(raw.pep)[1]
p <- dim(raw.pep)[2]
```

#Fit an outcome model (imputation)

We fit an outcome model using both low-dimensional covariates and entire peptide matrix with missingness.
A variant of VAE, called VAEIT, is employed. 

If you use your own dataset, change batches_cate and batches_cont accordingly.
batches_cate only accepts categrical variables, and batches_cont only accept continuous variables.

```{r fitting outcome}

batches_cate=NULL
batches_cont=matrix(as.matrix(covariate),ncol=ncol(covariate),byrow=F)
        
data <-raw.pep; data[is.na(data)] <- 0.;
mask <- - as.matrix(is.na(as.matrix(raw.pep)))
vae.impute<-vae(data,mask,batches_cate,batches_cont, NULL,dist_block=list('Gaussian'),
                        c(dim(data)[2]), num_epoch=300L, verbose=FALSE)[[1]]


```

## Construct pseudo-outcomes

Pseudo-outcomes are constructed through a predefined formula.
Logit model is employed to fit propensity scores, which is required to construct pseudo-outcomes.

```{r construct pseudo-outcomes}

fitd=raw.pep

  for(i.p in 1:p){
    mask=1-is.na(raw.pep[,i.p])
    prop=glm(mask~.,data=covariate, family = "binomial")
    fitd[,i.p]<-as.vector(prop$fitted.values)
  }

  AG=(raw.pep-vae.impute); AG[which(is.na(AG),arr.ind = T)]=0
  DR.pep=vae.impute+AG/fitd 

```

## Linear regression

We perform a linear regression for each peptide on low-dimensional covariates.The covariate of interest is selected from the first column of the covariate dataframe (subject to change for analysis purposes).

We utilize a heteroskedastic-consistent estimator to estimate the variance of the regression coefficient. Under the assumption of Missing Completely at Random (MCAR), this simplifies to using the usual Ordinary Least Squares (OLS) estimator.

Linear regression is fitted using doubly robust psuedo-outcomes (proposed) and the using the imputation outcomes (plugin method).  

```{r linear regression}


pval=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))
tval=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))
beta=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))

for(i.p in 1:p){
   fit=lm(DR.pep[,i.p]~., data=covariate)
    beta$DR_UW[i.p]=coef(summary(fit))[2,1] 
    tval$DR_UW[i.p]=beta$DR_UW[i.p]/sqrt(vcovHC(fit)[2,2])
    pval$DR_UW[i.p]=2*pt(abs(tval$DR_UW[i.p]), n-1, lower.tail=FALSE) 
    
    fit=lm(vae.impute[,i.p]~W1+W2, data=covariate)
    pval$plugin[i.p]=coef(summary(fit))[2,4] 
    tval$plugin[i.p]=coef(summary(fit))[2,3] 
    beta$plugin[i.p]=coef(summary(fit))[2,1] 

}


 
```

# Multiple Testing

We employ the p-values estimated from linear regressions to conduct a multiple testing procedure.
The p-values are transformed to Benjamini-Hochberg’s q-values, and we identify indices whose q-values are less than a predefined cutoff (tFDR).

```{r multiple testing, echo=F}

qval=data.frame(DR_UW=rep(0,p), plugin=rep(0,p))
qval$DR_UW=p.adjust(pval$DR_UW, method = "BH")
qval$plugin=p.adjust(pval$plugin, method = "BH")

tFDR=0.05
which(qval$DR_UW<tFDR);

```

# Evaluation

When synthetic were used, we know the ground truth of truely associated peptides (indicies in ``true.peptide"). Therefore, FDR and power can be computed and compared with the case when the plugin values are used. 

```{r evaluation}

power_DR=length(intersect(which(qval$DR_UW<tFDR),true.peptide))/length(true.peptide)
FDR_DR=length(setdiff(which(qval$DR_UW<tFDR),true.peptide))/length(which(qval$DR_UW<tFDR))
power_DR;FDR_DR

power_plugin=length(intersect(which(qval$plugin<tFDR),true.peptide))/length(true.peptide)
FDR_plugin=length(setdiff(which(qval$plugin<tFDR),true.peptide))/length(which(qval$plugin<tFDR))
power_plugin;FDR_plugin


```


